---
title: "Comprehensive Tutorial: Stochastic Reserving Models"
author: "R. Mark Sharp"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Comprehensive Tutorial: Stochastic Reserving Models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 7,
  fig.height = 5
)
```

## Introduction

This tutorial demonstrates all five stochastic reserving models implemented in
the `stochasticreserver` package:
1. **Chain Ladder** - Standard development factor approach
2. **Cape Cod** (Bornhuetter-Ferguson variant) - Exposure-based method
3. **Berquist-Sherman** - Incremental severity method with trend
4. **Hoerl Curve** - Smooth curve with shared operational time parameters
5. **Wright** - Generalized Hoerl with individual accident year levels

These methods are based on Roger Hayne's paper "A Flexible Framework for
Stochastic Reserving Models" published in *Variance* journal. The key insight
is that all five methods can be unified under a common maximum likelihood
estimation framework, enabling consistent uncertainty quantification.

## Theoretical Framework

### Uncertainty in Loss Reserving

Loss reserve estimates face three distinct sources of uncertainty:

1. **Process Uncertainty** - Inherent stochastic fluctuation in outcomes even
   when parameters are known exactly. This represents the irreducible randomness
   in insurance claim development.

2. **Parameter Uncertainty** - Error from estimating unknown parameters. Even
   with the correct model structure, finite data leads to estimation error.

3. **Model Uncertainty** - Risk that the assumed model structure doesn't match
   actual data-generating processes. Different reserving methods may yield
   different estimates, revealing where assumptions require investigation.

This package addresses process and parameter uncertainty through maximum
likelihood estimation (MLE). Model uncertainty is addressed by enabling
comparison across multiple methods.

### Maximum Likelihood Estimation

MLEs have desirable asymptotic properties (as sample size increases):

- **Consistency**: Estimates converge to true parameter values
- **Efficiency**: Variance achieves the theoretical lower bound (Cramér-Rao)
- **Asymptotic Normality**: Distribution approaches Gaussian with mean $\theta$
  and covariance equal to the inverse Fisher information matrix

The **Fisher Information Matrix** is:

$$I(\theta)_{ij} = E\left[\frac{\partial^2}{\partial\theta_i\partial\theta_j}
(-\ln L(X|\theta))\right]$$

This enables uncertainty quantification through the variance-covariance matrix
$\Sigma = I(\theta)^{-1}$.

### Stochastic Model for Incremental Averages

All models share a common likelihood structure. For incremental averages
$A_{ij}$ (where $i$ = accident year, $j$ = development lag), we assume:

$$A_{ij} \sim N(g_{ij}(\theta), \sigma^2_{ij})$$

The expected value $g_{ij}(\theta)$ is model-specific (Chain Ladder, Cape Cod,
etc.), while the variance structure is shared:

$$\sigma^2_{ij} = e^{\kappa - w_i} \cdot (g_{ij}(\theta)^2)^p$$

where:

- $\kappa$ is a proportionality constant
- $w_i = \log(d_i)$ where $d_i$ is the exposure count for accident year $i$
- $p$ is a power parameter controlling heteroscedasticity

### Negative Log-Likelihood Function

For a single observation, the negative log-likelihood is:

$$\ell(x; \mu, \kappa, p) = \frac{1}{2}\left(\kappa - w +
\ln(2\pi(\mu^2)^p)\right) + \frac{(x-\mu)^2}{2e^{\kappa-w}(\mu^2)^p}$$

The total objective function sums over all available cells in the triangle.

### The Five Model Forms

Each model specifies a different expected value function $g_{ij}(\theta)$:
| Model | Expected Value Form | Parameters |
|-------|---------------------|------------|
| **Chain Ladder** | $\theta_j \cdot U_i$ | Development proportions (sum to 1) |
| **Cape Cod** | $\theta_0 \cdot \theta_i^{row} \cdot \theta_j^{col}$ | Level + row/column factors |
| **Berquist-Sherman** | $\theta_i \cdot e^{\theta_{trend} \cdot j}$ | Year levels + trend |
| **Hoerl Curve** | $e^{\alpha + \beta_1\tau + \beta_2\tau^2 + \beta_3\ln(\tau) + \beta_4 i}$ | Curve params + row trend |
| **Wright** | $e^{\alpha_i + \beta_1\tau + \beta_2\tau^2 + \beta_3\ln(\tau)}$ | Individual levels + curve |

where $\tau$ represents operational time (development lag) and $U_i$ represents
the ultimate for accident year $i$.

### Gradient and Hessian

Efficient optimization requires analytical gradients and Hessians. For parameter
vector $\mathbf{a} = (\theta_1, \ldots, \theta_k, \kappa, p)$:

**Gradient with respect to model parameters:**
$$\frac{\partial \ell}{\partial \theta} = \sum_{i,j} g'_{ij}(\theta) \cdot
\left(\frac{p}{\mu_{ij}} + \frac{\mu_{ij} - A_{ij}}{\sigma^2_{ij}} -
\frac{p(A_{ij} - \mu_{ij})^2}{\sigma^2_{ij} \mu_{ij}}\right)$$

**Gradient with respect to variance parameters:**
$$\frac{\partial \ell}{\partial \kappa} = \frac{1}{2}\sum_{i,j}
\left(1 - \frac{(A_{ij} - \mu_{ij})^2}{\sigma^2_{ij}}\right)$$

The Hessian matrix enables Newton-type optimization and provides the information
matrix for uncertainty quantification.

```{r load-packages}
library(stochasticreserver)
library(mvtnorm)
library(abind)
library(stats)
```

## Data

### Package Data: Hayne's Example Triangle

The package includes data from the reference paper - a 10x10 development
triangle of cumulative averages (`B0`) and exposure counts (`dnom`).

```{r load-data}
# Load the cumulative average triangle
B0 <- stochasticreserver::B0
dnom <- stochasticreserver::dnom
size <- nrow(B0)

# Display the triangle
cat("Cumulative Average Triangle (B0):\n")
print(round(B0, 2))

cat("\nExposure Counts (dnom):\n")
print(round(dnom, 0))
```

### Data Preparation

We need to prepare several derived quantities for the models:

```{r data-prep}
# Calculate incremental average matrix (A0)
A0 <- cbind(B0[, 1], B0[, 2:size] - B0[, 1:(size - 1)])

# Generate log exposure matrix for variance structure
logd <- log(matrix(dnom, size, size))

# Create row and column index matrices
rowNum <- row(A0)
colNum <- col(A0)

# Create masks for data availability
# upper_triangle_mask: TRUE for cells with available data
upper_triangle_mask <- (size - rowNum) >= colNum - 1

# msn: mask for next forecast diagonal (one-period ahead)
msn <- (size - rowNum) == colNum - 2

# msd: mask for current diagonal (paid to date)
msd <- (size - rowNum) == colNum - 1

# Calculate amount paid to date for each accident year
paid_to_date <- rowSums(B0 * msd, na.rm = TRUE)

cat("Paid to Date by Accident Year:\n")
print(round(paid_to_date, 2))
```

### Simulated Data: Auto Liability Example

To demonstrate the models on different data characteristics, we also create a
simulated dataset representing a stylized auto liability triangle. This
simulated data exhibits:
- Moderate development over 10 periods
- Slight trend in accident year severity
- Realistic exposure variation

**Note:** This data is simulated. For production use, replace with actual
public data such as triangles from regulatory filings or industry studies.

```{r simulate-data}
set.seed(42)

# Simulate a second triangle with different characteristics
simulate_triangle <- function(n_years = 10, base_ult = 5000, trend = 0.03,
                              dev_pattern = NULL) {
  if (is.null(dev_pattern)) {
    # Typical auto liability development pattern
    dev_pattern <- c(0.35, 0.25, 0.15, 0.10, 0.06, 0.04, 0.025, 0.015,
                     0.008, 0.002)
    dev_pattern <- dev_pattern / sum(dev_pattern)  # Normalize
  }

  n_dev <- length(dev_pattern)

  # Generate ultimates with trend
  ultimates <- base_ult * (1 + trend)^(0:(n_years - 1))
  ultimates <- ultimates * exp(rnorm(n_years, 0, 0.05))  # Add noise

  # Generate incremental averages
  incr_avg <- matrix(NA, n_years, n_dev)
  for (i in 1:n_years) {
    for (j in 1:n_dev) {
      if ((n_years - i) >= (j - 1)) {
        mu <- ultimates[i] * dev_pattern[j]
        sigma <- mu * 0.1  # 10% coefficient of variation
        incr_avg[i, j] <- rnorm(1, mu, sigma)
      }
    }
  }

  # Convert to cumulative
  cum_avg <- t(apply(incr_avg, 1, cumsum))

  # Generate exposures (claim counts)
  exposures <- round(runif(n_years, 35000, 50000))

  list(
    B0 = cum_avg,
    A0 = incr_avg,
    dnom = exposures,
    ultimates = ultimates
  )
}

sim_data <- simulate_triangle()
B0_sim <- sim_data$B0
A0_sim <- sim_data$A0
dnom_sim <- sim_data$dnom

cat("Simulated Cumulative Triangle:\n")
print(round(B0_sim, 2))
```

## Helper Functions

We define helper functions to streamline the analysis across all models.

```{r helper-functions}
#' Fit a stochastic reserving model
#'
#' @param model_name Character: "Chain", "CapeCod", "Berquist", "Hoerl",
#'   or "Wright"
#' @param B0 Cumulative average matrix
#' @param A0 Incremental average matrix
#' @param dnom Exposure vector
#' @param paid_to_date Paid amounts to date
#' @param upper_triangle_mask Logical mask for available data
#' @return List with fitted model results
fit_model <- function(model_name, B0, A0, dnom, paid_to_date,
                      upper_triangle_mask) {

  size <- nrow(B0)
  logd <- log(matrix(dnom, size, size))

  # Select model
  model_lst <- switch(model_name,
    "Chain" = chain(B0, paid_to_date, upper_triangle_mask),
    "CapeCod" = capecod(B0, paid_to_date, upper_triangle_mask),
    "Berquist" = berquist(B0, paid_to_date, upper_triangle_mask),
    "Hoerl" = hoerl(B0, paid_to_date, upper_triangle_mask),
    "Wright" = wright(B0, paid_to_date, upper_triangle_mask)
  )

  g_obj <- model_lst$g_obj
  g_grad <- model_lst$g_grad
  g_hess <- model_lst$g_hess
  a0 <- model_lst$a0

  # Get starting values for kappa and p
  E <- g_obj(a0)
  yyy <- (A0 - E)^2
  yyy <- logd + log(((yyy != 0) * yyy) - (yyy == 0))
  sss <- na.omit(data.frame(x = c(log(E^2)), y = c(yyy)))
  if (nrow(sss) > 2) {
    ttt <- array(coef(lm(sss$y ~ sss$x)))[1:2]
  } else {
    ttt <- c(10, 1)
  }
  a0 <- c(a0, ttt)

  # Objective functions
  l.obj <- function(a, A) {
    make_negative_log_likelihood(a, A, dnom, g_obj)
  }

  l.grad <- function(a, A) {
    make_gradient_of_objective(a, A, dnom, g_obj, g_grad)
  }

  l.hess <- function(a, A) {
    make_log_hessian(a, A, dnom, g_obj, g_grad, g_hess)
  }

  # Minimize
  max_ctrl <- list(iter.max = 10000, eval.max = 10000)
  scale_vec <- abs(1 / (2 * ((a0 * (a0 != 0)) + (1 * (a0 == 0)))))

  mle <- tryCatch(
    nlminb(a0, l.obj, gradient = l.grad, hessian = l.hess,
           scale = scale_vec, A = A0, control = max_ctrl),
    error = function(e) {
      # Fallback without hessian
      nlminb(a0, l.obj, gradient = l.grad,
             scale = scale_vec, A = A0, control = max_ctrl)
    }
  )

  # Extract results
  npar <- length(mle$par) - 2
  p <- mle$par[npar + 2]
  mean_fitted <- g_obj(mle$par[1:npar])
  var_fitted <- exp(-outer(logd[, 1], rep(mle$par[npar + 1], size), "-")) *
    (mean_fitted^2)^p
  stres <- (A0 - mean_fitted) / sqrt(var_fitted)

  # Information matrix and variance-covariance
  hess_final <- l.hess(mle$par, A0)
  inf_mat <- tryCatch(
    solve(hess_final),
    error = function(e) NULL
  )

  list(
    model_name = model_name,
    mle = mle,
    npar = npar,
    mean = mean_fitted,
    var = var_fitted,
    stres = stres,
    g_obj = g_obj,
    vcov = inf_mat,
    logd = logd,
    convergence = mle$convergence
  )
}

#' Calculate forecast reserves
#'
#' @param fit Fitted model object
#' @param dnom Exposure vector
#' @param upper_triangle_mask Mask for available data
#' @return Data frame with reserve estimates
calculate_reserves <- function(fit, dnom, upper_triangle_mask) {
  size <- nrow(fit$mean)

  # Forecast for lower triangle (future payments)
  forecast_mask <- !upper_triangle_mask
  forecast_mean <- fit$mean * forecast_mask
  forecast_var <- fit$var * forecast_mask

  # Total reserve by accident year
  reserve_mean <- rowSums(dnom * forecast_mean, na.rm = TRUE)
  reserve_sd <- sqrt(rowSums(dnom^2 * forecast_var, na.rm = TRUE))

  # Aggregate
  total_mean <- sum(reserve_mean)
  total_sd <- sqrt(sum(dnom^2 * forecast_var, na.rm = TRUE))

  data.frame(
    accident_year = c(1:size, "Total"),
    reserve_mean = c(reserve_mean, total_mean),
    reserve_sd = c(reserve_sd, total_sd),
    cv = c(reserve_sd / pmax(reserve_mean, 1), total_sd / total_mean)
  )
}

#' Run Monte Carlo simulation
#'
#' @param fit Fitted model object
#' @param dnom Exposure vector
#' @param upper_triangle_mask Mask for available data
#' @param nsim Number of simulations
#' @return Matrix of simulated reserves
run_simulation <- function(fit, dnom, upper_triangle_mask, nsim = 5000) {
  if (is.null(fit$vcov)) return(NULL)

  size <- nrow(fit$mean)
  npar <- fit$npar

  # Masks for simulation
  smsk <- aperm(array(c(upper_triangle_mask), c(size, size, nsim)),
                c(3, 1, 2))

  # Sample parameters
  spar <- tryCatch(
    rmvnorm(nsim, fit$mle$par, fit$vcov),
    error = function(e) NULL
  )
  if (is.null(spar)) return(NULL)

  # Simulated means
  esim <- fit$g_obj(spar)

  # Simulated variances
  ksim <- exp(aperm(outer(array(spar[, npar + 1], c(nsim, size)),
                          log(dnom), "-"), c(1, 3, 2)))
  psim <- array(spar[, npar + 2], c(nsim, size, size))
  vsim <- ksim * (esim^2)^psim

  # Simulate future averages
  sim_avg <- array(rnorm(nsim * size * size, c(esim), sqrt(c(vsim))),
                   c(nsim, size, size))

  # Calculate reserves
  sdnm <- t(matrix(dnom, size, nsim))
  reserves <- sdnm * rowSums(sim_avg * !smsk, dims = 2)

  cbind(reserves, Total = rowSums(reserves))
}
```

## Model Fitting: All Five Methods

Now we fit all five models to the Hayne example data.

```{r fit-all-models}
model_names <- c("Chain", "CapeCod", "Berquist", "Hoerl", "Wright")

results <- list()
for (model_name in model_names) {
  cat("Fitting", model_name, "model...\n")
  results[[model_name]] <- fit_model(
    model_name, B0, A0, dnom, paid_to_date, upper_triangle_mask
  )
  cat("  Convergence:", results[[model_name]]$convergence,
      " Parameters:", results[[model_name]]$npar + 2, "\n")
}
```

## Model 1: Chain Ladder

The Chain Ladder model represents the standard actuarial approach where
development factors link successive development periods. In this stochastic
formulation, the expected emergence is parameterized such that development
proportions sum to 1.

### Parameters

```{r chain-params}
chain_fit <- results[["Chain"]]
cat("Chain Ladder Model\n")
cat("==================\n")
cat("Number of parameters:", chain_fit$npar + 2, "\n")
cat("  - Model parameters:", chain_fit$npar, "(development proportions)\n")
cat("  - Variance parameters: 2 (kappa, p)\n\n")

# Development pattern implied
npar <- chain_fit$npar
theta <- chain_fit$mle$par[1:npar]
dev_pattern <- c(theta, 1 - sum(theta))
cat("Implied development pattern:\n")
names(dev_pattern) <- paste0("Lag", 1:length(dev_pattern))
print(round(dev_pattern, 4))
cat("\nCumulative:", round(cumsum(dev_pattern), 4), "\n")
```

### Fitted Values

```{r chain-fitted}
cat("\nFitted Mean (expected incremental averages):\n")
print(round(chain_fit$mean, 2))
```

### Reserve Estimates

```{r chain-reserves}
chain_reserves <- calculate_reserves(chain_fit, dnom, upper_triangle_mask)
knitr::kable(chain_reserves, digits = 0,
             caption = "Chain Ladder Reserve Estimates")
```

## Model 2: Cape Cod (Bornhuetter-Ferguson)

The Cape Cod model incorporates exposure information through a multiplicative
structure with row (accident year) and column (development) factors. This
allows for prior information about ultimate severity levels.

### Parameters

```{r capecod-params}
cc_fit <- results[["CapeCod"]]
cat("Cape Cod Model\n")
cat("==============\n")
cat("Number of parameters:", cc_fit$npar + 2, "\n")
cat("  - Model parameters:", cc_fit$npar, "\n")
cat("    - 1 level parameter\n")
cat("    -", (size - 1), "row factors\n")
cat("    -", (size - 1), "column factors\n")
cat("  - Variance parameters: 2\n")
```

### Reserve Estimates

```{r capecod-reserves}
cc_reserves <- calculate_reserves(cc_fit, dnom, upper_triangle_mask)
knitr::kable(cc_reserves, digits = 0,
             caption = "Cape Cod Reserve Estimates")
```

## Model 3: Berquist-Sherman

The Berquist-Sherman model is an incremental severity method that explicitly
models trend over development periods. This is particularly useful when
development patterns are changing over time.

### Parameters

```{r berquist-params}
berq_fit <- results[["Berquist"]]
cat("Berquist-Sherman Model\n")
cat("======================\n")
cat("Number of parameters:", berq_fit$npar + 2, "\n")
cat("  - Model parameters:", berq_fit$npar, "\n")
cat("    -", size, "accident year averages\n")
cat("    - 1 trend parameter\n")
cat("  - Variance parameters: 2\n\n")

# Extract trend parameter
trend_param <- berq_fit$mle$par[size + 1]
cat("Estimated trend parameter:", round(trend_param, 4), "\n")
cat("This implies", round((exp(trend_param) - 1) * 100, 2),
    "% change per development period\n")
```

### Reserve Estimates

```{r berquist-reserves}
berq_reserves <- calculate_reserves(berq_fit, dnom, upper_triangle_mask)
knitr::kable(berq_reserves, digits = 0,
             caption = "Berquist-Sherman Reserve Estimates")
```

## Model 4: Hoerl Curve

The Hoerl model uses a smooth mathematical curve to describe development
rather than discrete factors. The curve includes polynomial and logarithmic
terms in operational time, plus a row trend.

### Mathematical Form

$$\mu_{ij} = \exp(\alpha + \beta_1 \tau_j + \beta_2 \tau_j^2 +
\beta_3 \log(\tau_j) + \beta_4 \cdot i)$$

where $\tau_j$ is the development lag (operational time).

### Parameters

```{r hoerl-params}
hoerl_fit <- results[["Hoerl"]]
cat("Hoerl Curve Model\n")
cat("=================\n")
cat("Number of parameters:", hoerl_fit$npar + 2, "\n")
cat("  - Model parameters:", hoerl_fit$npar, "\n")
cat("    - alpha (intercept)\n")
cat("    - beta1 (linear in tau)\n")
cat("    - beta2 (quadratic in tau)\n")
cat("    - beta3 (log tau)\n")
cat("    - beta4 (row trend)\n")
cat("  - Variance parameters: 2\n\n")

hoerl_theta <- hoerl_fit$mle$par[1:5]
names(hoerl_theta) <- c("alpha", "beta1", "beta2", "beta3", "beta4")
print(round(hoerl_theta, 4))
```

### Reserve Estimates

```{r hoerl-reserves}
hoerl_reserves <- calculate_reserves(hoerl_fit, dnom, upper_triangle_mask)
knitr::kable(hoerl_reserves, digits = 0,
             caption = "Hoerl Curve Reserve Estimates")
```

## Model 5: Wright (Generalized Hoerl)

The Wright model extends the Hoerl curve by allowing individual level
parameters for each accident year rather than just a linear trend. This
provides more flexibility at the cost of additional parameters.

### Mathematical Form

$$\mu_{ij} = \exp(\alpha_i + \beta_1 \tau_j + \beta_2 \tau_j^2 +
\beta_3 \log(\tau_j))$$

where $\alpha_i$ is an individual intercept for each accident year.

### Parameters

```{r wright-params}
wright_fit <- results[["Wright"]]
cat("Wright Model\n")
cat("============\n")
cat("Number of parameters:", wright_fit$npar + 2, "\n")
cat("  - Model parameters:", wright_fit$npar, "\n")
cat("    -", size, "individual year levels (alpha_i)\n")
cat("    - 3 operational time parameters (beta1, beta2, beta3)\n")
cat("  - Variance parameters: 2\n\n")

wright_theta <- wright_fit$mle$par[1:wright_fit$npar]
cat("Accident year levels:\n")
print(round(wright_theta[1:size], 4))
cat("\nOperational time parameters:\n")
names(wright_theta[(size + 1):(size + 3)]) <- c("beta1", "beta2", "beta3")
print(round(wright_theta[(size + 1):(size + 3)], 4))
```

### Reserve Estimates

```{r wright-reserves}
wright_reserves <- calculate_reserves(wright_fit, dnom, upper_triangle_mask)
knitr::kable(wright_reserves, digits = 0,
             caption = "Wright Model Reserve Estimates")
```

## Model Comparison

### Reserve Estimates Comparison

```{r compare-reserves}
comparison <- data.frame(
  Accident_Year = 1:size,
  Chain = chain_reserves$reserve_mean[1:size],
  CapeCod = cc_reserves$reserve_mean[1:size],
  Berquist = berq_reserves$reserve_mean[1:size],
  Hoerl = hoerl_reserves$reserve_mean[1:size],
  Wright = wright_reserves$reserve_mean[1:size]
)

# Add totals
comparison <- rbind(
  comparison,
  data.frame(
    Accident_Year = "Total",
    Chain = chain_reserves$reserve_mean[size + 1],
    CapeCod = cc_reserves$reserve_mean[size + 1],
    Berquist = berq_reserves$reserve_mean[size + 1],
    Hoerl = hoerl_reserves$reserve_mean[size + 1],
    Wright = wright_reserves$reserve_mean[size + 1]
  )
)

knitr::kable(comparison, digits = 0,
             caption = "Reserve Comparison Across Models")
```

### Parameter Count Comparison

```{r param-comparison}
param_df <- data.frame(
  Model = model_names,
  Model_Params = sapply(results, function(x) x$npar),
  Variance_Params = 2,
  Total_Params = sapply(results, function(x) x$npar + 2),
  Convergence = sapply(results, function(x) x$convergence)
)
knitr::kable(param_df, caption = "Parameter Count by Model")
```

### Negative Log-Likelihood Comparison

```{r nll-comparison}
nll_df <- data.frame(
  Model = model_names,
  NegLogLik = sapply(results, function(x) x$mle$objective),
  AIC = sapply(results, function(x) {
    2 * (x$npar + 2) + 2 * x$mle$objective
  }),
  BIC = sapply(results, function(x) {
    n_obs <- sum(!is.na(A0) & upper_triangle_mask)
    log(n_obs) * (x$npar + 2) + 2 * x$mle$objective
  })
)
nll_df <- nll_df[order(nll_df$AIC), ]
knitr::kable(nll_df, digits = 2, row.names = FALSE,
             caption = "Model Fit Statistics (sorted by AIC)")
```

## Diagnostic Plots

### Residual Analysis

Standardized residuals should be approximately standard normal if the model
fits well.

```{r residual-plots, fig.height=8}
par(mfrow = c(3, 2))

for (model_name in model_names) {
  fit <- results[[model_name]]
  stres <- fit$stres

  # Q-Q plot
  qqnorm(c(stres), main = paste(model_name, "- Q-Q Plot"))
  qqline(c(stres), col = "red")
}

# Combined residual comparison
par(mfrow = c(1, 1))
```

### Residuals by Calendar Year

```{r residual-by-cy, fig.height=6}
par(mfrow = c(2, 3))

for (model_name in model_names) {
  fit <- results[[model_name]]
  stres <- fit$stres
  cy <- rowNum + colNum - 1

  plot(c(cy), c(stres),
       main = paste(model_name, "- by Calendar Year"),
       xlab = "Calendar Year", ylab = "Std Residual",
       pch = 16, col = "blue")
  abline(h = 0, col = "gray")

  # Add mean by CY
  cy_means <- tapply(c(stres), c(cy), mean, na.rm = TRUE)
  lines(as.numeric(names(cy_means)), cy_means, col = "red", lwd = 2)
}

par(mfrow = c(1, 1))
```

### Residuals by Development Lag

```{r residual-by-lag, fig.height=6}
par(mfrow = c(2, 3))

for (model_name in model_names) {
  fit <- results[[model_name]]
  stres <- fit$stres

  plot(c(colNum), c(stres),
       main = paste(model_name, "- by Lag"),
       xlab = "Development Lag", ylab = "Std Residual",
       pch = 16, col = "darkgreen")
  abline(h = 0, col = "gray")

  # Add mean by lag
  lag_means <- colMeans(stres, na.rm = TRUE)
  lines(1:size, lag_means, col = "red", lwd = 2)
}

par(mfrow = c(1, 1))
```

## Monte Carlo Simulation

We run simulations to generate full reserve distributions for each model.

```{r run-simulations}
set.seed(123)
nsim <- 5000

sim_results <- list()
for (model_name in model_names) {
  cat("Running simulation for", model_name, "...\n")
  sim_results[[model_name]] <- run_simulation(
    results[[model_name]], dnom, upper_triangle_mask, nsim
  )
}
```

### Distribution of Total Reserves

```{r sim-distributions, fig.height=6}
par(mfrow = c(2, 3))

for (model_name in model_names) {
  sim <- sim_results[[model_name]]
  if (!is.null(sim)) {
    hist(sim[, "Total"], breaks = 50, main = model_name,
         xlab = "Total Reserve", col = "lightblue", border = "white")
    abline(v = mean(sim[, "Total"]), col = "red", lwd = 2)
    abline(v = quantile(sim[, "Total"], c(0.05, 0.95)),
           col = "blue", lwd = 2, lty = 2)
  }
}

par(mfrow = c(1, 1))
```

### Summary Statistics from Simulation

```{r sim-summary}
sim_summary <- data.frame(
  Model = character(),
  Mean = numeric(),
  SD = numeric(),
  CV = numeric(),
  Q05 = numeric(),
  Q50 = numeric(),
  Q95 = numeric(),
  stringsAsFactors = FALSE
)

for (model_name in model_names) {
  sim <- sim_results[[model_name]]
  if (!is.null(sim)) {
    total <- sim[, "Total"]
    sim_summary <- rbind(sim_summary, data.frame(
      Model = model_name,
      Mean = mean(total),
      SD = sd(total),
      CV = sd(total) / mean(total),
      Q05 = quantile(total, 0.05),
      Q50 = quantile(total, 0.50),
      Q95 = quantile(total, 0.95)
    ))
  }
}

knitr::kable(sim_summary, digits = 0, row.names = FALSE,
             caption = "Simulation Summary: Total Reserves")
```

## Application to Simulated Data

We now apply all models to the simulated auto liability data to demonstrate
robustness across different data characteristics.

```{r fit-simulated}
# Prepare simulated data
size_sim <- nrow(B0_sim)
logd_sim <- log(matrix(dnom_sim, size_sim, size_sim))
rowNum_sim <- row(A0_sim)
colNum_sim <- col(A0_sim)
upper_triangle_mask_sim <- (size_sim - rowNum_sim) >= colNum_sim - 1
msd_sim <- (size_sim - rowNum_sim) == colNum_sim - 1
paid_to_date_sim <- rowSums(B0_sim * msd_sim, na.rm = TRUE)

results_sim <- list()
for (model_name in model_names) {
  cat("Fitting", model_name, "to simulated data...\n")
  results_sim[[model_name]] <- fit_model(
    model_name, B0_sim, A0_sim, dnom_sim, paid_to_date_sim,
    upper_triangle_mask_sim
  )
}
```

### Comparison on Simulated Data

```{r compare-simulated}
comparison_sim <- data.frame(
  Model = model_names,
  Reserve = sapply(results_sim, function(x) {
    res <- calculate_reserves(x, dnom_sim, upper_triangle_mask_sim)
    res$reserve_mean[size_sim + 1]
  }),
  True_IBNR = sum(sim_data$ultimates * dnom_sim) -
    sum(B0_sim * upper_triangle_mask_sim * dnom_sim, na.rm = TRUE)
)
comparison_sim$Error_Pct <- (comparison_sim$Reserve -
  comparison_sim$True_IBNR) / comparison_sim$True_IBNR * 100

knitr::kable(comparison_sim, digits = c(0, 0, 0, 1), row.names = FALSE,
             caption = "Reserve Estimates vs True IBNR (Simulated Data)")
```

## Model Selection Guidance

Based on the analysis above, here is guidance for model selection:

### When to Use Each Model

1. **Chain Ladder**: Best for stable, mature lines with consistent development.
   Fewest parameters (9) provides robustness but less flexibility.

2. **Cape Cod**: Preferred when you have reliable prior information about
   ultimate severity levels. The separate row and column factors (19 parameters)
   provide flexibility but require more data to estimate reliably.

3. **Berquist-Sherman**: Use when you suspect systematic changes in development
   patterns over time. The explicit trend parameter (11 total) helps identify
   and quantify development year trends.

4. **Hoerl Curve**: Good for smooth development patterns. With only 5 model
   parameters, it's parsimonious while still capturing key features. The
   mathematical form ensures sensible extrapolation.

5. **Wright**: Most flexible individual year treatment (13 parameters). Use
   when accident years have genuinely different characteristics that shouldn't
   be constrained to a linear trend.

### Model Uncertainty

As Hayne emphasizes, practitioners should "use multiple methods" because
divergent forecasts indicate where underlying assumptions require investigation.
The differences between models reflect **model uncertainty** - a key component
of total reserve uncertainty often underappreciated in practice.

## Conclusion

This tutorial demonstrated all five stochastic reserving models in the
`stochasticreserver` package:

- **Chain Ladder**: Simplest, most constrained
- **Cape Cod**: Exposure-based with separate factors
- **Berquist-Sherman**: Explicit development trend
- **Hoerl**: Smooth parametric curve
- **Wright**: Individual year levels

The unified maximum likelihood framework enables:
1. Consistent parameter estimation across models
2. Proper uncertainty quantification
3. Model comparison via information criteria
4. Monte Carlo simulation for reserve distributions

For production use, consider:
- Running multiple models and comparing results
- Examining residual diagnostics carefully
- Using simulation to capture parameter uncertainty
- Recognizing that model differences indicate structural uncertainty

## References

Hayne, R. "A Flexible Framework for Stochastic Reserving Models."
*Variance*, Vol 7, Issue 2. Available at:
<https://variancejournal.org/article/120823>

## Session Information

```{r session-info}
sessionInfo()
```
